Unsupervised embedding of trajectories captures the latent structure of mobility

### 2020-10-24 paper/current/main.tex
main text, yy comments

Human mobility and migration drive major societal phenomena such as the growth and evolution of cities~\autocite{wef2017migration, curiel2018citygravity},  epidemics~\autocite{kraemer2020covid, truscott2012epidemicgravity}, economies~\autocite{kaluza2010cargo, kerr2011immigration}, and innovation~\autocite{kaiser2018innovation, sugimoto2017mostimpact, petersen2018multiscale}.
Historically, human mobility has been strongly constrained by physical separation---geographic distance~\autocite{zipf1946gravity}.
However, geographic distance is becoming less relevant in the increasingly-globalized world~\autocite{czaika2018globalisation, buchan2009globalization} in which physical barriers are shrinking while linguistic, cultural, and historical relationships are becoming more important.
As understanding mobility is becoming critical for contemporary society, finding frameworks that can capture this complexity are of paramount importance.
Here, using three datasets that capture human trajectories,  we demonstrate that a neural embedding model~\autocite{mikolov2013word2vec} can encode nuanced relationships between locations into a vector-space, providing an effective measure of distance that reflects the multi-faceted structure of human mobility. 
Focusing on the case of scientific mobility, we show that embeddings of scientific organizations uncovers cultural and linguistic relations at multiple levels of granularity, and even academic prestige.
Furthermore, the embedding vectors reveal universal relationships between organizational characteristics, and their place in the global landscape of scientific mobility. 
The ability to learn scalable, dense and meaningful representations of mobility directly from the data can open up a new avenue of studying mobility across domains.

@@

where $m_{i}$ is the population of locations $i$,  $f(r_{ij})$ is a decay function with respect to distance between locations, and $C$ is a constant estimated from data (see Methods). 
For the flight itinerary data, we use population $m_{i}$ as the total number of unique passengers on each airport, and for scientific mobility, we use the mean annual number of unique mobile and non-mobile authors who were affiliated with each organization.
$\hat{T}_{ij}$, which is often referred to as ``expected flux''~\autocite{simini2012universal}, is the expected frequency of the co-occurrence of location $i$ and $j$ in the trajectory in the gravity model.
The gravity model dictates that the expected flow, $\hat{T}_{ij}$, is symmetric (\ie $\hat{T}_{ij }= \hat{T}_{ji})$), is proportional to the locations' population, $\hat{T}_{ij} \propto m_{i} m_{j}$, and decays as a function of their distance, $f(r_{ij})$. 
We define the distance function in terms of either the geographic distance between locations or their functional distance in the vector space, which is calculated as the cosine distance between their vectors, termed the \textit{embedding distance}. 
For geographic distance, we define $f(r_{ij})$ as the standard power-law function, and for the embedding distance, we use the exponential function, selected as the best performing for each case (Fig.~\ref{fig:supp:predict_emb_bymodel} and Fig.~\ref{fig:supp:predict_geo_bymodel}).

@@

In the remaidner of the paper, we focus only on scientific mobility, leveraging its richness to investigate how the geometric space generated by the neural embedding method sheds light on the multi-faceted relationships between organizations. 
To explore the topological structure of the embedding, we use a topology-based dimensionality reduction method (UMAP~\autocite{mcinnes2018umap}) to obtain a two-dimensional representation of the embedding space (Fig.~\ref{fig:projection}a).
By showing relationships between individual organizations, rather than aggregates such as nations or cities, this projection constitutes the largest and highest resolution ``map'' of scientific mobility to date.

@@

We report that this curve is almost universal across many countries. 
For instance, China's curve closely mirrors that of the United States (Fig.~\ref{fig:length}b).
Smaller but scientifically advanced countries such as Australia, and other populous countries such as Brazil also exhibit curves similar to the United States (Fig.~\ref{fig:length}b, inset).
Other nations exhibit different curves which lacks the portions with decreasing norm, probably indicating the lack of highly prestigious institutions.
Similar patterns can be found across many of the 30 of the countries with the most total researchers (Fig.~\ref{fig:concave30}) see \SI{} for more discussion).

>>>>

Human mobility and migration drive major societal phenomena such as the growth and evolution of cities~\autocite{wef2017migration, curiel2018citygravity},  epidemics~\autocite{kraemer2020covid, truscott2012epidemicgravity}, economies~\autocite{kaluza2010cargo, kerr2011immigration}, and innovation~\autocite{kaiser2018innovation, sugimoto2017mostimpact, petersen2018multiscale}.
Historically, human mobility has been strongly constrained by physical separation---geographic distance~\autocite{zipf1946gravity}.
However, geographic distance is becoming less relevant in the increasingly-globalized world~\autocite{czaika2018globalisation, buchan2009globalization} in which physical barriers are shrinking while linguistic, cultural, and historical relationships are becoming more important.
As understanding mobility is becoming critical for contemporary society, finding frameworks that can capture this complexity are of paramount importance.
Here, using three discrete human trajectory datasets,  we demonstrate that a neural embedding model~\autocite{mikolov2013word2vec} can encode nuanced relationships between locations into a vector-space, providing an effective measure of distance that reflects the multi-faceted structure of human mobility. 
Focusing on the case of scientific mobility, we show that embeddings of scientific organizations uncovers cultural and linguistic relations at multiple levels of granularity, and even academic prestige.
Furthermore, the embedding vectors reveal universal relationships between organizational characteristics, and their place in the global landscape of scientific mobility. 
The ability to learn scalable, dense and meaningful representations of mobility directly from the data can open up a new avenue of studying mobility across domains.

@@

where $m_{i}$ is the population of locations $i$,  $f(r_{ij})$ is a decay function with respect to distance between locations, and $C$ is a constant estimated from data (see Methods). 
For the flight itinerary data, we use population $m_{i}$ as the total number of unique passengers on each airport, and for scientific mobility, we use the mean annual number of unique mobile and non-mobile authors who were affiliated with each organization.
$\hat{T}_{ij}$, which is often referred to as ``expected flux''~\autocite{simini2012universal}, is the expected frequency of the co-occurrence of location $i$ and $j$ in the trajectory in the gravity model.
The gravity model dictates that the expected flow, $\hat{T}_{ij}$, ($\hat{T}_{ij }= \hat{T}_{ji}$), is proportional to the locations' population, $\hat{T}_{ij} \propto m_{i} m_{j}$, and decays as a function of their distance, $f(r_{ij})$. 
We define the distance function in terms of either the geographic distance between locations or their functional distance in the vector space, which is calculated as the cosine distance between their vectors, termed the \textit{embedding distance}. 
For geographic distance, we define $f(r_{ij})$ as the standard power-law function, and for the embedding distance, we use the exponential function, selected as the best performing for each case (Fig.~\ref{fig:supp:predict_emb_bymodel} and Fig.~\ref{fig:supp:predict_geo_bymodel}).

@@

In the remainder of the paper, we focus on scientific mobility, leveraging its richness to investigate how the geometric space generated by the neural embedding method sheds light on the multi-faceted relationships between organizations. 
To explore the topological structure of the embedding, we use a topology-based dimensionality reduction method (UMAP~\autocite{mcinnes2018umap}) to obtain a two-dimensional representation of the embedding space (Fig.~\ref{fig:projection}a).
By showing relationships between individual organizations, rather than aggregates such as nations or cities, this projection constitutes the largest and highest resolution ``map'' of scientific mobility to date.

@@

We report that this curve is almost universal across many countries. 
For instance, China's curve closely mirrors that of the United States (Fig.~\ref{fig:length}b).
Smaller but scientifically advanced countries such as Australia, and other populous countries such as Brazil also exhibit curves similar to the United States (Fig.~\ref{fig:length}b, inset).
Other nations exhibit different curves which lacks the portions with decreasing norm, probably indicating the lack of internationally-prestigious institutions.
Similar patterns can be found across many of the 30 of the countries with the most total researchers (Fig.~\ref{fig:concave30}; see \SI{} for more discussion).
